The low-frequency Murchison radio environment:\\interference analysis and mitigation

### 2014-10-27 articles/2014/MWA-Radio-Environment/mwa-radio-environment.tex
Started on the conclusions

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of the latter approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. Analysing the results of interference detection and improving the understanding of RFI contamination can aid in several ways: understanding the effects on the science; improved observation scheduling; designing hardware and location of future telescopes with a maximal cost/benefit approach; and enforcing spectrum management strategies that are most beneficial for the observatories.

@@

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI mitigation effectivity. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose any issues on regular observations despite that, unlike the MWA, LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.

@@

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern -- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. This flagger has shown to have a good accuracy, is fast and has a library interface, which allows it to be integrated in a pipeline.

@@

The \textsc{aoflagger} software provides a C++ library%TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see TODO fix reference Ord et al. (2014). Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is then directly readable by the astronomical software.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. The \textsc{ndppp} software used by LOFAR (TODO reference?) uses a similar partitioning of data to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers might have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations.

@@

\section{Data description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The observation nights are listed in Table~\ref{tbl:obs-list}.

@@

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the number of removed edge channels to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DVB, the EoR high band is again slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy. TODO what are these transmitters?

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM and 2-m amateur bands show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and because of RFI leakage, but because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

@@

Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive for steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

@@

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor. The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively.

@@

Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz). This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all. This variability might be caused by intrinsic variation or movement of the transmitter or varying propagation conditions. While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot. In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is showed but is zoomed in on a briefly-observed RFI event at 150.17 MHz. This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; TODO cite) might help to analyze these phenomena.

Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation. The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter is not completely detected (Fig.~\ref{fig:156_7}, bottom-left plot). The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot). This kind of detection is currently not implemented in \textsc{cotter}.

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. On a high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, approximately 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

\section{Discussion}
\begin{itemize}
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
\end{itemize}

\noindent\begin{figure}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9cm]{img/LOFAR-MWA-occupancy}
\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are regridded to the same frequency resolution of 48~kHz. }
\label{fig:lofar-mwa-occupancy}
\end{center}
\end{figure}

\subsection{Comparison with LOFAR}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes RFI that is only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations. Taking into account these differences, it becomes hard to compare the environments fairly based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible is thrown away but the detection is sufficient for further data reduction. For the MWA, this goal was reached by throwing away a factor of two less data.

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. While the difference between a 1.65\% or 3.18\% loss of data is hardly a reason to spend resources to build at a remote site like the Murchison Radio Observatory, there are additional reasons to do so. A benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the FM-station bands (88--108~MHz). Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. Such a detection requires very deep integrations, and it could be that RFI becomes problematic when reaching lower noise levels. However, initial results with LOFAR predict that RFI will not prevent a detection of EoR signals \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions}
\begin{itemize}
 \item Much less RFI than LOFAR (but at a cost)
 \item About half the observations don't need flagging at all. 
 \item FM not completely empty
 \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
\end{itemize}

>>>>

While some success has been seen in filtering interference from observations in which the data samples are cleaned from the contribution of interference \citep{spatial-filtering-parkes-multibeam-for-pulses,rfi-spatial-processing-hellbourg-2014}, such an approach is often not feasible due to technical limitations or the type of RFI. Therefore, a common approach is detection of contaminated data and ignoring these samples in further data analysis \citep{post-correlation-rfi-classification}. The consequence of this approach is that a certain amount of data is lost due to interference, and that frequencies that are continuously occupied by transmitters can not be observed. It is important to analyze the impact of RFI on a specific instrument to consider the right RFI mitigation approach. Additionally, an increased understanding of the RFI situation will help in several other ways: understanding the effects of RFI on the science; observation scheduling; designing robust hardware; chosing locations of future telescopes with a maximal cost/benefit approach; and designing effective spectrum management strategies.

@@

The interferometric arrays LOFAR, PAPER and the Long Wavelength Array (LWA) are all designed to perform 21-cm EoR science, and all observe in approximately the same frequency range as the MWA. Comparing the MWA to these arrays therefore allows the comparison of the observatory site and hardware design with respect to the RFI impact. TODO say something about PAPER and MWA RFI. In \citet{lofar-radio-environment} it has been shown that the radio environment of LOFAR does not pose issues on regular observations even though LOFAR is build in a populated area. This is confirmed by \citet{ncp-eor-yatawatta}, where the authors analyse the first imaging of EoR long-integration observations with LOFAR. They reach near-thermal noise sensitivity, and argue that RFI is not a limitation for further increasing the sensitivity. Further analysis showed that RFI models for LOFAR can estimate the leakage of the RFI detection \citep{offringa-rfi-distributions}. Those analyses concluded that with sufficient precautions, such as good receiver design and accurate detection methods, the leaked RFI is weak and averages down similar to Gaussian noise.

@@

\section{Method}
RFI detection, often referred to as ``data flagging'', is one of the first steps in the data processing. It is important to perform initial data flagging at high time and frequency resolution, because this increases detection accuracy and decreases the loss of data \citep{lofar-radio-environment}. Consequently, RFI detection has to work on large data volumes, and its computational cost is therefore a concern --- in particular for many-element arrays. In some projects, simple thresholding is used to mitigate the worst interference, which is computationally cheap but not very accurate. For example, a $3\sigma$ threshold is used for reducing PAPER data in \citet{parsons-paper-eorlimit-2014}. Several observatories or projects have designed pipelines that include more advanced RFI mitigation. Examples of such pipelines include \textsc{aoflagger} \citep{post-correlation-rfi-classification,scale-invariant-rank-operator}, originally designed for LOFAR; \textsc{flagcal} for preprocessing data from the Giant Metrewave Radio Telescope (GMRT; \citealt{prasad-flagcal-2012}); and \textsc{serpent} for preprocessing data from the Multi-Element Radio Linked Interferometer Network (e-MERLIN; \citealt{serpent-peck-2013}). For RFI detection in MWA observations, LOFAR's \textsc{aoflagger} is used. It has been shown that this flagger has a good accuracy and is fast. It also has a library interface, which allows it to be integrated in a pipeline.

@@

The \textsc{aoflagger} software provides a C++ library %TODO \footnote{\url{http://aoflagger.sourceforge.net/doc/api/}}
that can be integrated in a pipeline, thereby minimizing reading/writing of data. We have written an MWA-specific preprocessing pipeline named \textsc{cotter} that uses the \textsc{aoflagger} library for RFI detection. RFI detection is only performed on cross-correlations. Auto-correlations are normally removed, because they are not used in imaging. In addition to the RFI detection, \textsc{cotter} performs the following steps: it converts the raw correlator files into \textsc{casa} measurement sets or uvfits files; applies bandpass gain corrections; corrects the phases for varying cable lengths; calculates the $uvw$-coordinates; applies phase tracking to the desired sky coordinates; flags samples from the correlator that are missing or incorrect; and allows averaging the observation in frequency or time direction to reduce the data volume. It also collects various statistics and writes these into a measurement set using the LOFAR quality statistics format\footnote{Described in ``MeasurementSet description for LOFAR version 2.08'' by Schoenmakers \& Renting.}. Various tools are available to analyse these statistics, e.g. the \textsc{aoqplot} tool that is part of the \textsc{aoflagger} software can plot the statistics over various dimensions in an interactive manner.

MWA observations are split into snapshots of a few minutes by the correlator. The MWA archive stores the raw correlator outputs for every snapshot as an observation that can be referenced by its observation ID. For more details about the correlator, see TODO fix reference Ord et al. (2014). Currently, the \textsc{cotter} preprocessing pipeline is ran by the scientist that calibrates and images the data. Once the scientist has downloaded the raw files for a given observation ID, there are various ways of processing MWA data. For imaging MWA data with the Real-Time System (RTS; \citealt{rts-mwa-2008}), \textsc{cotter} is ran in a special mode such that it only flags the data, and does not convert the raw correlator files. After running \textsc{cotter}, the raw files and a flag mask are given as input to the RTS. For more traditional data processing with e.g. \textsc{casa} or \textsc{miriad}, the \textsc{cotter} output is set to either the \textsc{casa} or uvfits format. The output is then directly readable by the most common astronomical software packages.

One particular issue in implementing \textsc{cotter} is that both the raw correlator files and the desired output files are ordered in time, but flagging is done baseline by baseline. The \textsc{sumthreshold} technique and the scale-invariant rank operator that are used by the flagging strategy, use statistics over large time and frequency ranges. Therefore, accuracy is improved when the time-frequency flagging intervals are as large as possible. However, a typical snapshot is about 50 GB in size, and without increasing reading and writing, it requires thus 50 GB of memory to perform flagging on the full data. When less memory is available, \textsc{cotter} will split the observation in different time chunks and flags those independently. This is similar to the partitioning that is used in the \textsc{ndppp} software used by LOFAR (TODO reference?) to overcome the time-ordering problem. Partitioning the data has the undesirable behaviour that executing \textsc{cotter} on a low-memory machine decreases its flagging accuracy. Because not all astronomers have easily access to high-memory machines, a platform with sufficient memory has been setup that pre-runs \textsc{cotter} on all observations. For this use-case, \textsc{cotter} is run in a flagging-only mode on the high-memory machine. The astronomer downloads the raw files and the flagging output files, and reruns \textsc{cotter} to convert the raw files and apply the RFI detection from the first run.

@@

\section{Data description}
We analyze the RFI in 10 nights of data from two different MWA projects: the MWA EoR project \citep{bowman-science-with-the-mwa-2013} and the GaLactic and ExtrAgalactic MWA (GLEAM) survey (TODO cite Randall). The nights of observations are listed in Table~\ref{tbl:obs-list}.

@@

The poly-phase filter bank that performs the first separation into subbands, introduces a 1.28~MHz periodic spectral signature. This shape is e.g. visible in Fig.~\ref{fig:bandpass}. The poly-phase filter band-pass shape is hard to take out completely, because the filters are imperfect and each sub-band is effected by leakage from signals in its adjacent sub-bands. The leakage from adjacent sub-bands manifests itself as if the sub-band band-pass is direction dependent. To solve this, the bordering 80~kHz on both sides of a 1.28~Mhz subband are normally flagged by \textsc{cotter}. This implies that in normal observations, 13\% of the data is lost due to the poly-phase filter. However, while 80~kHz is sufficient to prevent imaging artefacts, we noticed that the RFI statistics were still slightly biased by the edge channels. Therefore, for the analyses in this paper we increase the removed bandwidth to 200~kHz on either side of each 1.28~MHz sub-band, or 32\% in total.

@@

Fig.~\ref{fig:rfi-per-band} shows the overal detected RFI occupancy per sub-band, as calculated over all observations except GLEAM 2014-03-17. The latter is the only night affected by interference from digital video broadcasting (DVB), and will be analysed later. The FM bands around 100~MHz and the ORBCOMM bands around 138~MHz are clearly present in the data. Excluding the RFI from DVB, the EoR high band is slightly cleaner than the EoR low band, as its worst subband at 188.2 MHz has 1.03\% occupancy. The sub-bands at 145.9 and 149.8~MHz in the EoR low band have both 2.1\% occupancy. TODO what are these transmitters?

The residual noise levels after flagging can be used to validate whether the flagged data is free of RFI. In Fig.~\ref{fig:stddev-per-set}, the residual noise levels are plotted per high-resolution channel for each of the observations. Observation `GLEAM 2014-03-17' shows residual DVB interference, and it is clear that this RFI has not been adequately flagged. Therefore, observations have to be cleaned of DVB with another method. Additionally, some channels in the FM radio band show standard deviations that deviate slightly as a result from the smaller amount of available data after flagging and possibly because of RFI leakage. Nevertheless, because the effect is small these frequencies can be calibrated and imaged without a problem. The subband at 137~MHz that is occupied by ORBCOMM is hard to calibrate because of the small amount of residual data per channel, and possibly also because of residual RFI. With the exception of the ORBCOMM frequencies and DVB affected nights, the RFI detection employed in \textsc{cotter} is sufficient to allow calibration and imaging without further RFI mitigation efforts. This is verified by early imaging results from the GLEAM survey and the MWA commissioning survey \citep{mwacs-2014}.

@@

Because significant DVB interference is left in `GLEAM 2014-03-17' after flagging, we have analysed this night more extensively. The DVB transmitters are terrestrial, and the fact that only this night observes the DVB, implies the RFI must originate from an over-the-horizon transmitter that is reflected by unusually strong ionospheric activity. As can be seen in the left plot of Fig.~\ref{fig:dvb}, this kind of RFI can fully contaminate frequencies 174--195~MHz, thus over half of the instantaneous bandwidth. Because the \textsc{aoflagger} determines its thresholding levels from the data, and because it needs to be insensitive to steep RMS jumps over frequency due to the varying coarse channel gains (see Fig.~\ref{fig:bandpass}), it is insensitive for RFI that is this broad.

@@

The distribution curves of RFI-detected samples do not show a very significant power-law tail. Fits over regions selected by eye have been added to Fig.~\ref{fig:lognlogs}. The EoR high-band observes too little RFI to generate such a power law, which is reflected by its fit of $N=\beta S^{-4.47}$, with $\beta$ being a scaling factor. The GLEAM and EoR low-band observations do show a small power-law part, but the fitted exponent depends strongly on what part of the tail is selected. The selected regions result in $N=\beta S^{-1.43}$ and $N=\beta S^{-1.33}$ for the GLEAM and Eor-low curves, respectively. Due to the small part over which the power laws hold, as well as the high possible error in the fits because a fitting region needs to be selected, it is hard to infer if a uniform distribution of transmitters is observed.

@@

Fig.~\ref{fig:2m-amateur-band} shows two examples of RFI events in the same EoR low-band snapshot: the top panels display the Stokes~I values of a single correlation, in which a transmitter has been observed in the 2-m amateur band (146 MHz). This is the worst contamination in our data sets by this transmitter, and there are also snapshots in which the transmitter is not visible at all. This variability might be caused by intrinsic variation, movement of the transmitter or varying propagation conditions. While a change in pointing can also change the appearance of the transmitter, the beam does not change within a single snapshot, while in Fig.~\ref{fig:2m-amateur-band} the strength and affected bandwidth of the transmitter does change during the snapshot. In the bottom panel of Fig.~\ref{fig:2m-amateur-band}, the same snapshot is showed but is zoomed in on a briefly-observed RFI event at 150.17 MHz. This event occupies only a single 40~kHz channel for 2 seconds, and thus is an event which requires flagging at approximately the observed time and frequency resolution or higher for accurate detection.

Besides persistent transmissions that occupy a few channels, transient broad-band events are observed as well. Occassionally, DVB RFI is visible for a brief moment, as for example visible in Fig.~\ref{fig:dvb-burst}. As can be seen in the right plot of Fig.~\ref{fig:dvb-burst}, such a brief interference is well flagged by the \textsc{aoflagger}. Another transient broad-band example is displayed in Fig.~\ref{fig:broadband-pulse}, which shows strong broad-band pulses of a second in length. Strong pulses such as these are rare and well flagged, but a few weak broad-band pulses are observed in almost every 2-min snapshot. These weak pulses are not visible nor detected in a single baseline correlation, but can be seen in a dynamic spectrum when the power of all baselines are added together. We currently do not know what their origin is, but it is likely they are \textit{in situ} generated RFI or receiver saturations much shorter than the integration time. The MWA Voltage Capture System (VCS; TODO cite), which allows high time resolution observations, might help to analyze these phenomena.

Fig.~\ref{fig:156_7} shows a longer RFI contamination at 156.66~MHz that is hardly visible in a single correlation. The \textsc{aoflagger} detects this RFI partially (Fig.~\ref{fig:156_7}, top-right plot), but when plotting the standard deviation over all correlations in a dynamic spectrum, it is evident that this RFI transmitter extends beyond what is detected (Fig.~\ref{fig:156_7}, bottom-left plot). The detection becomes more complete when the \textsc{aoflagger} is executed on the standard deviations over all baselines (Fig.~\ref{fig:156_7}, bottom-right plot). This kind of detection is currently not implemented in \textsc{cotter}.

\subsection{Computational performance of \textsc{cotter}}
Because \textsc{cotter} processes the data at high resolution, its computational performance is an important consideration. A major contribution to the runtime is the reading and writing of the data, and the runtime is thus influenced by the input-output (IO) performance of the host system. Excluding IO, the main computational performance of \textsc{cotter} consists of running \textsc{aoflagger} on the data and collecting the statistics; performing all its other tasks such as applying the cable delays and averaging increases the runtime by approximately a percent. A high-end desktop with 32 GB of memory and a 3.20-GHz Intel Core i7-3930K processor with six cores, and with a 5 disc RAID setup, is used to time \textsc{cotter}. The wall-clock runtime for processing a single 2~min snapshot with a 0.5~s / 40~kHz input resolution, using common averaging settings to output at 2~s / 80~kHz resolution, is split up as follows: 3~min are spend on reading the data, 5~min on RFI detection and 3.5~min on writing the data. Real-time processing can therefore be achieved by using 6 of such nodes in parallel.

\noindent\begin{figure}
\begin{center}\hspace*{-0.2cm}\includegraphics[width=9cm]{img/LOFAR-MWA-occupancy}
\caption{Comparison between LOFAR and MWA RFI occupancies. The statistics are resampled to the same frequency resolution of 48~kHz. }
\label{fig:lofar-mwa-occupancy}
\end{center}
\end{figure}

\section{Comparison with LOFAR}
The frequency range of the LOFAR high-band antennas (HBAs) overlaps with the MWA frequency range, and this therefore allows to compare the RFI situations at the same frequencies. The RFI occupancies for LOFAR from \citet{lofar-radio-environment} and for the MWA from this work are plotted over the HBA frequency range of 115--163~MHz in Fig.~\ref{fig:lofar-mwa-occupancy}. The statistics are regridded to the same frequency resolution of 48~kHz. The average RFI occupancy in this frequency range is 1.65\% for the MWA, while O2012 reports a 3.18\% occupancy for LOFAR.

It should be noted that the RFI detection for LOFAR was performed at 0.78~kHz while for the MWA it is performed at 40~kHz. As shown in O2012, LOFAR observes many RFI events that are only a single or a few 0.78~kHz channels wide, and consequently due to MWA's lower frequency resolution, the MWA will only detect the brightest of such transmissions. O2012 also show that ionospheric scintillation of Cassiopeia~A triggers detection events, and this is one of the reasons for LOFAR's relatively high minimum occupancy level of $\sim2\%$ RFI, in comparison to 0.5\% for the MWA. In the MWA data, no false detections have been seen that are caused by ionospheric scintillation, likely because of the absence of the strongest sources: Cassiopeia~A is never visible at the MWA, and Cygnus~A is only visible at low declinations. An additional explanation could also be that the sidelobe behaviour between MWA and LOFAR is also different due to the difference between MWA tiles and LOFAR stations. Because of these differences, it is hard to compare the environments based on detected RFI levels. Nevertheless, for both telescopes, the automatic detection strategies have been optimized such that as little data as possible are thrown away, but to be sufficient for further data reduction.

After RFI detection, power spectra from both MWA and LOFAR show smooth curves without significant leakage, and the data are clean enough for further processing and imaging. From the difference between 1.65\% or 3.18\% loss of data at the MWA and LOFAR sites respectively, it is clear that the impact of RFI is smaller for the MWA, but RFI mitigation is still required with similar computational requirements as for LOFAR. Both RFI occupancy levels are so small that the RFI contamination in the 115--163~MHz range should not be a consideration in selecting a site. However, a benefit of building the MWA at a remote radio-quiet site, is that the MWA can observe in the 88--108~MHz FM-station bands. Initial MWA experiments have confirmed the availability of these bands for science \citep{mckinley-moon-2013,mwacs-2014}. Moreover, for both LOFAR and MWA, a primary science driver is the detection of signals from the EoR. The required integration time for such a detection of approximately a hundred nights has currently not been achieved, and it could be that RFI becomes problematic when reaching lower noise levels. Initial results with LOFAR predict that RFI will not prevent such a detection \citep{ncp-eor-yatawatta, offringa-rfi-distributions}.

\section{Conclusions \& Discussion}
We have described the automated RFI detection strategy for the MWA and shown RFI statistics at various frequencies and over various nights. After RFI detection, data can be calibrated and imaged without a problem, with the exception of the ORBCOMM bands at 137~MHz. Also, DVD signals are seen $\sim3\%$ of the time, during which observing in the 174--195~MHz range produces unusable data. Over the full GLEAM range of 72--231~MHz, 1.1\% of the data is flagged by the RFI detection, and these RFI events are attributed to several different transmitters. Some residual RFI is seen in the FM-station bands, but these frequencies are usable after our described automated RFI detection. The issue of RFI has become smaller by building at a remote site, but still requires adequate handling.

\begin{itemize}
 \item Frequency behaviour of instrument makes things harder! Need to fix pass-band correction.
 \item High time resolution
 \item Location of SKA is not free of RFI --> Need good ADCs etc
 \item SKA -- how to implement a Cotter for SKA?
 \item PAPER etc needs to use combinatorial/morphological flagging
 \item The presence of strong interference implies there might be weak interference present that is not easily seen or detected. 
\end{itemize}
